---
title: "Best Practices with R"
author: "Juliano Palacios Abrantes"
date: '2019-10-16'
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_depth: 1
  pdf_document:
    toc: yes
    toc_depth: 1
---

# The main points from [Wilson *et al* 2017](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005510)

## Data management
- Save the raw data and backed up in more than one location.
- Create the data you wish to see in the world and analysis-friendly data.
- Record all the steps used to process data.
- Anticipate the need to use multiple tables, and use a unique identifier for every record.
- Submit (*finalized*) data to a reputable DOI-issuing repository so that others can access and cite it.

## Software
- Decompose programs into functions.
- Be ruthless about eliminating duplication.
- Always search for well-maintained software libraries that do what you need.
- Test libraries before relying on them.
- Provide a simple example or test data set.
- Submit code to a reputable DOI-issuing repository.

## Collaboration
- *Use version control*
- Create an overview of your project.
- Create a shared "to-do" list for the project.
- Decide on communication strategies (*e.g SLACK, Github, etc.*)
- Make the license explicit.
- Make the project citable.pipipipipipipipiGGGREAT!viijjjj

## Project organization
- Put each project in its own directory, which is named after the project. (*R projects!*)
- Put text documents associated with the project in the *doc* directory.
- Put raw data and metadata in a data directory and files generated during cleanup and analysis in a *results* directory.
- Put project source code in the *src* directory.
- Put external scripts or compiled programs in the *bin* directory.
- Name all files to reflect their content or function.

## Keeping track of changes
- Back up (almost) everything created by a human being as soon as it is created.
- Keep changes small.
- Share changes frequently.
- Create, maintain, and use a checklist for saving and sharing changes to the project.
- Use a version control system (*Github*).

## Manuscripts
- Write manuscripts using online tools with rich formatting, change tracking, and reference management.
- Write the manuscript in a plain text format that permits version control.

# Main points from me

## Some **general** points

- Always start with a clean environment instead of saving the workspace
- Use Rmarkdown instead of normal script
- Use version control even if its only you
- Comment smart! (*ProTip: shortcut for "#" **cmd + shift + c***)

## Keep a good and consistent **notation**

As you know there are many ways you can write in R, but there are good, bad, and ugly ways! 

- Use "<-" instead of "=" for naming variables (*ProTip:There's a shortcut for <-; alt - *)
- Forget the `setwd()`
  - Use projects 
  - And the `here` package when collaborating
- Be clear and keep a consistency in your variable naming
  - No spaces!
  - Caps?
  - Short and to the point

## The Good

```{r Notation_Good, eval = F, echo = T}

######----------------------------- READ ME ---------------------------#####
# This is code to give the "Best Practices wit R" workshop for the FOL group
# It was developed by Juliano Palacios on Oct. 2019
######-----------------------------------------------------------------#####


# NEVER use "=" for setting a variable, instead use "<-"
Pi_Value <- pi

# Forget the `setwd()` ####

# Get used to projects and folders, keep everything consistent and you'll have an easier coding life
# If collaborating use the "here" package

# Keep a consistency in your variable naming


Pi_Sqr <- pi^2
Cars_Data <- cars
Cars_Over_Pi <- subset(Cars_Data, Cars_Data$speed > Pi_Sqr)

# or... all low caps

pi_sqr <- pi^2
cars_data <- cars
cars_over_pi <- subset(cars_data, cars_data$speed > cars_data)
```

## The Bad
```{r Notation_Bad, eval = F, echo = T}

# Using "=" will lead to problems in advanced R skills (e.g. functions)
Pi_Value = pi

# Forget the `setwd()` ####
# setwd() to a path only you have
setwd("~\Users\juliano\path\that\only\I\have")


# Avoid names that are common case functions or meaningless
Cars <- cars 
mean <- mean(cars$speed)
sd <- sd(Cars$dist)
thisisaverylongvariablenameforsomethingsoshort <- pi
T <- t.test(Cars$speed, mu = thisisaverylongvariablenameforsomethingsoshort)
```

## The Ugly
```{r Notation_Ugly, eval = F, echo = T}

# Ughhh don't mix them! 
Pi_Value = pi
Two_pi <- Pi_Value^2

# Forget the `setwd()` ####
# setwd() for each path you have..
setwd("~/Users/juliano/Documents/path/that/only/I/have/Data")
setwd("~/Users/juliano/Documents/path/that/only/I/have/Save/Figures")
setwd("~/Users/juliano/Documents/path/that/only/I/have/Code")


# Keep a consistency in your variable naming
Cars_data <- cars
subsetcars_one <- subset(cars_data, cars_data$speed > cars_data)
subsetcars_two <- subset(cars_data, cars_data$speed < cars_data)
# .
# .
# .
subsetcars_n <- subset(cars_data, cars_data$speed < cars_data)

PI_sqr <- pi^2
Cars_data <- cars
subsetcars <- subset(cars_data, cars_data$speed > cars_data)
thisisaverylongvariablenameforsomethingsoshort <- pi
```

## Keep **track** of who wrote the code and its intended purpose

Commenting is very important and it can be the difference between spending hours trying to understand what you did versus just re-take wherever you left. 

*Beware that excess commenting could be as bad as no commenting!*

## The Good

```{r Keepig_track_Good, eval = F, echo = T}
# The hack function

#### ------------------------------- Read Me------------------------------- ###
# This function is used to hack the mclapply function so it works on Windows.
# It was created by Nathan VanHondus and accessed from R-Bloggers
# https://www.r-bloggers.com/implementing-mclapply-on-windows-a-primer-on-embarrassingly-parallel-computation-on-multicore-systems-with-r/
#### ---------------------------------------------------------------------- ###

# The Function
Mclapply_Hack <- function(...){
  
  ## Create a cluster
  size.of.list <- length(list(...)[[1]])
  
  cl <- makeCluster(min(size.of.list, n_cores)) # n_cores is the number of cores in the pc
  
  ## Find out the names of the loaded packages 
  loaded.package.names <- c(
    ## Base packages
    sessionInfo()$basePkgs,
    ## Additional packages
    names(sessionInfo()$otherPkgs))
  
  tryCatch( { # in case ther's an error
    
    ## Copy over all of the objects within scope to
    ## all clusters. 
    this.env <- environment()
    while( identical( this.env, globalenv() ) == FALSE ) {
      clusterExport(cl,
                    ls(all.names=TRUE, env=this.env),
                    envir=this.env)
      this.env <- parent.env(environment())
    }
    clusterExport(cl,
                  ls(all.names=TRUE, env=globalenv()),
                  envir=globalenv())
    
    ## Load the libraries on all the clusters
    ## N.B. length(cl) returns the number of clusters
    parLapply( cl, 1:length(cl), function(xx){
      lapply(loaded.package.names, function(yy) {
        require(yy , character.only=TRUE)})
    })
    
    ## Run the lapply in parallel 
    return( parLapply( cl, ...) )
  }, finally = {  # close try_catch
    ## Stop the cluster
    stopCluster(cl)
  })
  
  
```

## The Bad

```{r Keepig_track_Bad, eval = F, echo = T}

This_Function_I_Need <- function(...){
    size.of.list <- length(list(...)[[1]])
    cl <- makeCluster(min(size.of.list, n_cores))
    loaded.package.names <- c(sessionInfo()$basePkgs,names(sessionInfo()$otherPkgs))
    tryCatch( {
       this.env <- environment()
       while( identical( this.env, globalenv() ) == FALSE ) {clusterExport(cl, ls(all.names=TRUE, env=this.env),envir=this.env)
this.env <- parent.env(environment())}
       clusterExport(cl,ls(all.names=TRUE, env=globalenv()),envir=globalenv())
       parLapply( cl, 1:length(cl), function(xx){
           lapply(loaded.package.names, function(yy) {
               require(yy , character.only=TRUE)})
       })
       return( parLapply( cl, ...) )
    }, finally = {        
       stopCluster(cl)
    })
```

## The Ugly

```{r Keepig_track_Ugly, eval = F, echo = T}

#### ------------------------------- ####
# The hack function ####
#### ------------------------------- ####

#### ------------------------------- Read Me------------------------------- ###
# This function is used to hack the mclapply function so it works on Windows.
# It was created by Nathan VanHondus and accessed from R-Bloggers
# https://www.r-bloggers.com/implementing-mclapply-on-windows-a-primer-on-embarrassingly-parallel-computation-on-multicore-systems-with-r/
#### ---------------------------------------------------------------------- ###

#### ------------------------------- ####
# The Function
#### ------------------------------- ####

Mclapply_Hack <- function(...){
  #### ------------------------------- ####
  ## Create a cluster
  #### ------------------------------- ####
  size.of.list <- length(list(...)[[1]])
  # n_cores is the number of cores in the pc
  cl <- makeCluster(min(size.of.list, n_cores)) # get the minimum size of all clusters
  #### ------------------------------- ####
  ## Find out the names of the loaded packages 
  #### ------------------------------- ####
  loaded.package.names <- c(
    ## Base packages
    sessionInfo()$basePkgs,
    ## Additional packages
    names(sessionInfo()$otherPkgs))
  #### ------------------------------- ####
  # try catch moment
  #### ------------------------------- ####
  tryCatch( { # in case ther's an error
    #### ------------------------------- ####
    ## Copy over all of the objects within scope to
    ## all clusters. 
    #### ------------------------------- ####
    this.env <- environment()
    while( identical( this.env, globalenv() ) == FALSE ) {
      clusterExport(cl, #cluster of cores
                    ls(all.names=TRUE, env=this.env), # list them all
                    envir=this.env) # in this environent
      this.env <- parent.env(environment())
    }
    clusterExport(cl,#cluster of cores
                  ls(all.names=TRUE, env=globalenv()),# list them all
                  envir=globalenv()) # in this environment
    #### ------------------------------- ####
    ## Load the libraries on all the clusters
    ## N.B. length(cl) returns the number of clusters
    #### ------------------------------- ####
    parLapply( cl, 1:length(cl), function(xx){ # the parallel lapply
      lapply(loaded.package.names, function(yy) { # lapply it all!!
        require(yy , character.only=TRUE)})
    })
    #### ------------------------------- ####
    ## Run the lapply in parallel 
    #### ------------------------------- ####
    return( parLapply( cl, ...) )
  }, finally = {  # close try_catch
    ## Stop the cluster
    stopCluster(cl)
  })
  #### ------------------------------- #### ------------------------------- ####
  #### ---------------END OF THE ANALYSIS---------------- ####
#### ------------------------------- #### ------------------------------- ####
```

## *Libraries* can get messy quite quickly 

I like to have all of them at the beginning and load them as a first step in my code. However, I do know some people that prefer to load them as they go. I think this is up to you.

## The Good

```{r Libraries_Good, eval = F, echo = T}
# Libraries
library(readxl) # Read dataframe
library(tidyverse) # Data manipulation
library(rgdal) # Spatial analysis 

## Library hack!
# This function was developed by someone who I don't know (and badly did not get the reference) but it loads all packages and installs those you don't have.
# This is the first line of code in ALL of my R scripts

ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE,repos = "http://cran.us.r-project.org")
  sapply(pkg, require, character.only = TRUE)
}

packages <- c(
  "readxl", # Read dataframe
  "dplyr", # Data manipulation
  "tidyr" # Da
)
ipak(packages)
  
```

## The bad

```{r Libraries_Bad, eval = F, echo = T }
# Avoid loading them in the console or in the right menu --------------------------------------->
# This will require you to manually load them every time you open the script
# Also, not having explanation for new libraries could be a problem...
library(reshape) # ????
library(vegan) # ????

```

## The Ugly

```{r Libraries_Ugly, eval = F, echo = T}
# The ugly
# For me, not having all the things in one first chunk or part of the code sucks...
Cars_Data <- cars
library(ggplot2)
ggplot(Cars_Data) +
  geom_point(aes(speed,dist))

Pi_Sqr <- pi^2
Subset_Data <- dplyr::filter(Cars_Data, speed >= Pi_Sqr)
library(dplyr)
Subset_Data <- select(Cars_Data, speed)

```

input_file <- "data/data.csv" 
output_file <- "data/results.csv"


## Convert **paths** into variables

Not only paths but numbers or other variables that you will use a lot, its better to convert them into variables, that way you will reduce chances of carrying a mistake into the analysis

```{r Paths_Good, eval = F, echo = T}

Data_Path <- "./Data/"
Results_Path <- "./Results/"

# read data
Data_A <- read.csv(paste(Data_Path,"Data_A.csv",sep=""))
Data_B <- read.csv(paste(Data_Path,"Data_B.csv",sep=""))
Data_C <- read.csv(paste(Data_Path,"Data_C.csv",sep=""))

# Do some type of analysis
Final_Data <- rbind(Data_A,Data_B,Data_C)

Result <- some_analysis_function(Final_Data)

# write results
write.table(results,
            paste(Results_Path,"results.csv",sep="")
            )

# The same applies for regular numbers

# Let's say we want to know how many records we have of each species in the iris dataset that have a sepal width larger than 3

# Global variables
Spp <- unique(iris$Species)
Sepal_W <- 3

# My analysis

# Step one, filter the species
# Step two, filter spp with sepal w > 3
# Step three, estimate individuals left

# Setosa
Setosa <- subset(iris, iris$Species == Spp[1]) 
Setosa_Subset <- subset(Setosa, Setosa$Sepal.Width >= Sepal_W)
Setosa_Large <- nrow(Setosa_Subset)

# Versicolor
Versicolor <- subset(iris, iris$Species == Spp[2])
Versicolor_Subset <- subset(Versicolor, Versicolor$Sepal.Width >= Sepal_W)
Versicolor_Large <- nrow(Versicolor_Subset)

# Virginica
Virginica <- subset(iris, iris$Species == unique(iris$Species)[3])
Virginica_Subset <- subset(Virginica, Virginica$Sepal.Width >= Sepal_W)
Virginica_Subset <- nrow(Virginica_Subset)

### ------ NOTE ------- ###
# There is an even better way to do this... Next week! 
### ------------------- ###
```


```{r Paths_Bad, eval = F, echo = T}

# Read data
Data_A <- read.csv("./Data/Data_A.csv",sep="")
Data_B <- read.csv("./Data/Data_B.csv",sep="")
Data_C <- read.csv("./Data/Data_C.csv",sep="")

# Do some type of analysis
Final_Data <- rbind(Data_A,Data_B,Data_C)

Result <- some_analysis_function(Final_Data)

# Write results
write.table(results, "./Results/results.csv")

# The same applies for regular numbers

# Get the name of species in the iris dataset
unique(iris$Species)

# Setosa
Setosa <- subset(iris, iris$Species == unique(iris$Species)[1])
Setosa_Subset <- subset(Setosa, Setosa$Sepal.Width >= 3)
Setosa_Large <- nrow(Setosa_Subset)

# Versicolor
Versicolor <- subset(iris, iris$Species == unique(iris$Species)[2])
Versicolor_Subset <- subset(Versicolor, Versicolor$Sepal.Width >= 2)
Versicolor_Large <- nrow(Versicolor_Subset)

# Virginica
Virginica <- subset(iris, iris$Species == unique(iris$Species)[3])
Virginica_Subset <- subset(Virginica, Virginica$Sepal.Width >= 3)
Virginica_Subset <- nrow(Virginica_Subset)
```

